---
title: "Analysis of Large Wildland Fires in California, Nevada, and Idaho"
author: "Li Jia Go, Melissa Merritt, and Nanditha Satagopan"
date: "2022-12-14"
output:
  html_document: default
editor_options:
  chunk_output_type: console
---
Git Repository: https://github.com/lijiago/EDAProject_Go_Merritt_Satagopan
---
\newpage
\tableofcontents 
\newpage
\listoftables 

**Table 1**: Summary of Data Structure


\newpage
\listoffigures 

**Figure 1**: Figure showing histogram of large wildfires in 1988 and 2000, California

**Figure 2**: Map of location of large wildfires in California in 1988 and 2000

**Figure 3**: Large Wildfires in U.S. States Between 1984 and 2018

**Figure 4**: Mean Wildfire Size for California, Idaho, and Nevada from 1984 through 2018

**Figure 5**: Burnt Wildfire Area from 1988-2007 in California

**Figure 6**: Burnt Prescribed Fire Area from 1988-2007 in California

**Figure 7**: Burnt Wildland Use Fire Area from 1988-2007 in California

**Figure 8**: Pairwise Relationships between Mean Fire Sizes in California, Idaho, and Nevada

**Figure 9**: Time Series Analysis Plot of Wildfire Area from 1988-2007 in California

\newpage

```{r setup, include=FALSE}
#Set working directory

getwd()

#Load packages

library(dplyr)
library(scales)
library(formatR)
library(agricolae)
library(tidyverse)
library(mapview)
library(sf)
library(lubridate)
library(ggplot2)
install.packages("kableExtra")
library(kableExtra)

# Setting ggplot theme

mytheme <- theme_bw(base_size = 10) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top", 
        axis.text.x=element_text(angle = 60, hjust = 1)) 
theme_set(mytheme)

# Loading dataset

Fire_data <- read.csv("./Data/Processed/US_FireData_Processed.csv")

```


# Rationale and Research Questions

With climate change, recent observations show an increase in occurrences, severity and extent of wildfires across the United States. As disturbance regimes across the country change, we thought it was important to assess trends, differences, and changes in specific fire regimes. The analysis is aimed at trying to infer any potential differences and trends in the sizes and occurrences of large wildfires through the last few decades. Monitoring any trend in the large fires can help gain an understanding on the health of the landscape and perhaps predicting future incidents. The use of prescribed and wildland fires through the years also gives us an understanding of how they have gained popularity in landscape management over time.  

We decided to identify and focus our study on the three states with the most large fire occurrences (California, Idaho, and Nevada) because that would provide us with a larger dataset and better accuracy to derive conclusions.

**Research Questions:**

**Question 1:** 
Do the mean sizes of large wildfires in California differ between 1988 and 2000? 

**Question 2:** 
Is there a difference between large wildfire sizes in the 3 states with the most large wildfire occurrences? (California, Idaho, and Nevada) 

**Question 3:** 
Is there a trend in large fire occurrences in California from 1984 to 2007? What is the trend in wildfire, prescribed fires and wildland fire use over the same years? 


\newpage

# Dataset Information

  The dataset we used in our analysis is titled "Monitoring Trends in Burn Severity Burned Area Boundaries" (https://www.mtbs.gov/). The United States Forest Service published this data and it was first published on May 4th, 2017, but has since been modified on August 8th, 2022. This data was collected using the Landsat Thematic Mapper and USFS data from each fire incident. This data is a feature layer, so it is geospatial data. This dataset was a part of a project called "Monitoring Trends in Burn Severity (MTBS)" that "assesses the frequency, extent, and magnitude (size and severity) of all large wildland fires." The term large wildland fire includes wildfire, wildland use fires (fires set as a means of suppression), and prescribed burns. The size of a large wildland fire varies; in the western US, a large wildland fire is any fire larger than 1,000 acres; in the eastern US, a large wildland fire is any fire larger than 500 acres. This data covers the entire United States and includes Puerto Rico. 

  To wrangle the data, we started by reading the csv file into our rmarkdown document. We then were able to use the pipe function (%>%) with the select() function for the data set to import only the columns we needed for our analysis. These columns included the fire size (acres), fire id, state, fire name, year, start month, start day, fire type, latitude, and longitude. After we had the columns we needed, the next step was to use the substr() function on the fire id column because all we needed from this column was the state abbreviation. Using substr(), we were able to include only the first two characters in the ID. Once we only had the state abbreviations in this column, we renamed the column using the pipe function (%>%) and the rename() function to rename it to STATE. The final step was to use write.csv() to save the processed dataset to our processed folder. 
  

**Table 1**: Summary of Data Structure - This table describes what each column in our wrangle dataset represents. In the data range or description column, the quantitative data has ranges and the qualitative data has descriptions. 

```{r, echo=TRUE, warning=FALSE, message=FALSE}
dataset_info <- data.frame(
  Variable = c("Fire Size", "State", "Fire Name", "Year", "Start Month", "Start Day", "Fire Type", "Latitude", "Longitude"), 
  Units = c("Acres", "-", "-", "Year #", "Month #", "Day #", "-", "Degrees North", "Degrees West"), 
  Data_Range_or_Description = c("500 - 593,364", "U.S.A. Including Puerto Rico", "On Scene Fire Name", "Fire state year", "Fire start month number", "Fire start day number", "Wildfire, Wildland Use Fire, Prescribed Burn", "17.957 - 69.465", "-66.129 - -165.881"),
  Data_Type = c("Quantitative", "Qualitative","Qualitative", "Qualitative", "Qualitative", "Qualitative", "Qualitative", "Quantitative", "Quantitative")
)

kbl(dataset_info)  %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

\newpage

# Exploratory Analysis 

**Question 1 Exploratory Analysis:** 
We first filtered the dataset to only include wildfires in California for years 1988 and 2000. We then created a histogram of wildfires in 1988 & 2000 to explore the data. However, from the plot, we could not tell if the mean was the same between both of the chosen years. As such, for the actual analysis, we decided to run a two-sample t-test to test whether the mean of acreage of fires between the two years was equivalent. 


```{r, echo=TRUE, warning=FALSE, message=FALSE}
#filtering dataset for wildfires in California in years 1988 and 2000
California.Wildfires <- Fire_data %>%
  filter(STATE %in% "CA") %>%
  filter(FIRE_TYPE %in% "Wildfire") %>%
  filter(YEAR %in% c("1988", "2000"))

#data exploration
```

**Figure 1**:  Figure showing histogram of large wildfires in 1988 and 2000, California
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(California.Wildfires, aes(x = ACRES, color =as.factor(YEAR))) +
  geom_freqpoly() +
  labs(x='Acreage of fires', y='Count', color='Year', 
       title="Histogram of wildfires in 1988 & 2000, California")  


```

We also created a map to show the locations of each of the fires for both the years within the state. The purple icons represent fires that burned in 1988 while the yellow icons represent fires that burned in 2000.


```{r, echo=TRUE, warning=FALSE, message=FALSE}
#exploratory spatial plots 
California.Wildfires.sf <- California.Wildfires %>% 
  st_as_sf(coords=c('LONGITUDE', 'LATITUDE'), crs=4269)
```

**Figure 2**: Map of location of large wildfires in California in 1988 and 2000
```{r, echo=TRUE, warning=FALSE, message=FALSE}
mapview(California.Wildfires.sf, zcol="YEAR")
```


**Question 2 Exploratory Analysis:** 
We first filtered the dataset to only include wildfires and not wildland use fires or prescribed burns. In order to answer the question that looked at the states with the most fire occurrences, we used figures to see which states had the most large wildfire occurrences (Figure 3). We were able to determine that California, Nevada, and Idaho were the states with the most large wildfires, so we then filtered the dataset to only include those states. After filtering the dataset we wanted to see the trends of the mean large wildfire sizes for each year from 1984 to 2018 in these three states to give an idea of how the mean fire sizes compare to each other (Figure 4).


```{r, echo=TRUE, warning=FALSE, message=FALSE}
# We want to only include wildfires, so we need to wrangle the data using the filter() function. 

Wildfire_data <- Fire_data %>%
  filter(FIRE_TYPE == "Wildfire")

# We want to figure out which two states have the most wildfires

Count_per_state <- Wildfire_data %>%
  count(STATE)

# To view the amount of large wildfires for each space we can use ggplot() and within ggplot, the geom_bar() function.
```

**Figure 3**: Figure showing the number of large wildfires in each state with documented large wildfires from 1984-2000. 
```{r, echo=TRUE, warning=FALSE, message=FALSE}
Wildfires_state_plot <- ggplot(Wildfire_data) + 
  geom_bar(aes(x=STATE)) +
  theme(axis.text.x=element_text(angle = 60, hjust = 1)) +
  xlab("State") +
  ylab("Number of Large Wildfires Between 1984-2018") +
  labs(title = "Number of Large Wildfires in U.S. States Between 1984 and 2018" )
print(Wildfires_state_plot)

```
 
```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Now knowing that California, Idaho, and Nevada have the most wildfire occurrences, we can filter the data to only include July and August fires for those three states using the pipe function %>%. 

States_Wildfire_data <- Wildfire_data %>%
  filter(STATE == "CA" | STATE == "NV" | STATE == "ID")

# Now we need to wrangle the data, to get the mean wildfire sizes (acres) for the states with the most occurrences. 

Mean_wildfire_size_state <- States_Wildfire_data %>%
  group_by(STATE, YEAR) %>%
  summarise(Mean_wildfire_size = mean(ACRES)) 

# Using the mean wildfire dataset, we can plot the mean fire sizes data for each year. This will give us the line graph of each states mean wildfire sizes and how they have changed over the years.
```

**Figure 4**: Figure showing the trends in mean wildfire occurrences from 1984 to 2018 in California, Idaho, and Nevada.
```{r, echo=TRUE, warning=FALSE, message=FALSE}
mean_firesize_plot <- ggplot(Mean_wildfire_size_state, aes(x = YEAR, y = Mean_wildfire_size, color = STATE)) +
  geom_line() + 
  xlab("Year") +
  ylab("Mean Wildfire Size (Acres)") +
  labs(title = "Mean Wildfire Size for California, Idaho, and Nevada from 1984 through 2018")
print(mean_firesize_plot)

```


**Question 3 Exploratory Analysis:** 
We first filtered the dataset to only include wildfires in California for the years 1988 to 2007. Subsequent pipe functions were used to obtain different datasets for the three fire categories- wildfire, prescribed and wildland use fire to check usability of data. Since the data collection was unevenly distributed over the months of each year, mean fire area was calculated for each fire for better comparison. The plots for the fires showed different trends but it was apparent from the wrangled datasets that time series analyses could not be performed for prescribed and wildland use fires since there were large gaps in data.


```{r, echo=TRUE, warning=FALSE, message=FALSE}
summary(Fire_data)

#Wrangling dataset for California fires only
all_cal_fires <- Fire_data %>%
  filter(STATE=="CA")
summary(all_cal_fires)

cal_fires <- all_cal_fires %>%
  filter(FIRE_TYPE=="Wildfire" | FIRE_TYPE=="Prescribed Fire" | FIRE_TYPE=="Wildland Fire Use")
summary(cal_fires)

#Wrangling Wildfire dataset

wildfires_ca <- cal_fires %>%
  filter(FIRE_TYPE=="Wildfire")

wildfires_mean <- 
  wildfires_ca %>%
  group_by(YEAR) %>%
  summarize(Mean_fires = mean(ACRES))

#Plotting Wildfire Fires 1984-2007
```

**Figure 5**: Burnt Wildfire Area from 1988-2007 in California
```{r, echo=TRUE, warning=FALSE, message=FALSE}
wildfire_plot <- 
  ggplot(wildfires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Wildfire Area",
       y="Burnt Area in Acres",
       x="Years") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildfire_plot)

```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#Wrangling Prescribed fire dataset

prescribed_fires_ca <- cal_fires %>%
  filter(FIRE_TYPE=="Prescribed Fire")

prescribed_fires_mean <- 
  prescribed_fires_ca %>%
  group_by(YEAR) %>%
  summarize(Mean_fires = mean(ACRES)) 

#Plotting Prescribed Fires 1984-2007
```

**Figure 6**: Burnt Prescribed Fire Area from 1988-2007 in California
```{r, echo=TRUE, warning=FALSE, message=FALSE}
prescribed_fire_plot <- 
  ggplot(prescribed_fires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Prescribed Fire Area",
       y="Burnt Area in Acres",
       x="Years") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(prescribed_fire_plot)

```

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#Wrangling Wildland Use Fire Dataset

wildland_fires_ca <- cal_fires %>%
filter(FIRE_TYPE=="Wildland Fire Use")

wildland_fires_mean <- 
  wildland_fires_ca %>%
  group_by(YEAR) %>%
  summarize(Mean_fires = mean(ACRES))

#Plotting Wildland Fires 1984-2007
```
**Figure 7**: Burnt Wildland Use Fire Area from 1988-2007 in California
```{r, echo=TRUE, warning=FALSE, message=FALSE}
wildland_fire_plot <-
  ggplot(wildland_fires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Burnt Wildland Use Fire Area",
       y="Burnt Area in Acres",
       x="Years")+
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildland_fire_plot)

##We reject the the Prescribed and Wildland Fire data sets. Their time series cannot be plotted since there are huge gaps and great inconsistency in data and thus a significant trend or pattern cannot be deduced. 

```


\newpage

# Analysis



## Question 1: Do the mean sizes of large wildfires in California differ between 1988 and 2000?
To test whether the mean sizes of large wildfires in California differ between 1988 and 2000, we ran a t-sample t test to test the hypothesis. 

H0 (null hypothesis): Means between both years are equal
H1 (alternative hypothesis): Means between both years are not equal

From the two-sample test, we obtained a p-value of 0.3754, which was more than 0.05. This meant that we accept the null hypothesis that the means were statistically equal for both years. 

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#running two sample t test and printing the result of the test 
Twosamp.ttest <- t.test(California.Wildfires$ACRES ~ California.Wildfires$YEAR)
Twosamp.ttest

```

## Question 2: Is there a difference between large wildfire sizes in the 3 states with the most large wildfire occurrences? (California, Idaho, and Nevada) 
To test if there are differences in the large wildfire sizes in these three states, we ran an analysis of variance using an ANOVA test. 

H0 (null hypothesis): There is no difference between large wildfire sizes in the 3 states with the most large wildfire occurrences (California, Idaho, and Nevada). 
H1 (alternative hypothesis): There is a difference between large wildfire sizes in the 3 states with the most large wildfire occurrences (California, Idaho, and Nevada).

We first ran an ANOVA, and the result of our ANOVA was a p-value of 0.00274, which would lead us to reject the null hypothesis. The next test we ran was the linear model, which gave us the results that California and Nevada are statistically different (p-value = 0.000617), but Idaho is not. This is also confirmed in the Tukey HSD test, which you can see the results of in Figure 8. 

```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Run the first ANOVA using the aov() function

FireSize_ANOVA <- aov(data = States_Wildfire_data, ACRES ~ STATE)
summary(FireSize_ANOVA)

# Run the same ANOVA using the the lm() function to get more details

FireSize_ANOVA2 <- lm(data = States_Wildfire_data, ACRES ~ STATE)
summary(FireSize_ANOVA2)

# Use the HSD.test to extract the groupings for pairwise relationships. 

FireSize_groups <- HSD.test(FireSize_ANOVA, "STATE", group = TRUE)
FireSize_groups

# Plot the groupings using ggplot() 
```
**Figure 8**: Pairwise relationships between California, Idaho, and Nevada wildfire sizes. Groupings found using the Tukey's HSD test.
```{r, echo=TRUE, warning=FALSE, message=FALSE}
FireSize_plot <- ggplot(States_Wildfire_data, aes(x = STATE, y = ACRES)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("b", "ab", "a")) +
  labs(x = "State", y = "Fire Size (Acres)", title = "Pairwise Relationships between Mean Fire Sizes in California, Idaho, and Nevada") +
  ylim(0, 20000) 
print(FireSize_plot)
```


## Question 3:  Is there a trend in large fire occurrences in California from 1984 to 2007? What is the trend in wildfire, prescribed fires and wildland fire use over the same years? 

A time series analysis is done for wildfire by plotting mean burnt area vs. years to observe any potential trend over the years. Additionally, the Mann-Kendall test is performed to check for the nature of the possible trend.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
#Wildfire Time Series
#The time series analysis of the occurrences of large wildfires helps us infer if they is any seasonality and/or trend in the variables over time (here, from years 1984-2007).

wildfire_ts <- ts(wildfires_mean$Mean_fires, start = c(1984), frequency = 1)
plot(wildfire_ts)


#Plotting the Wildfire Time Series
```
**Figure 9**: Time Series Analysis Plot of Wildfire Area from 1988-2007 in California
```{r, echo=TRUE, warning=FALSE, message=FALSE}
wildfire_TS_plot <- 
  ggplot(wildfires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Wildfire Time Series",
       y="Burnt Area in Acres",
       x="Years 1984-2007") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildfire_TS_plot)

#Prescribed Fire Time Series
prescribed_fire_ts <- ts(prescribed_fires_mean$Mean_fires, start = c(1984), frequency = 1)

#Wildland Fire Time Series
wildland_fire_ts <- ts(wildland_fires_mean$Mean_fires, start = c(1984), frequency = 1)
```

**Mann-Kendall Test- Trend Analysis**

Running the Mann-Kendall Test to assess whether a set of data values is increasing over time or decreasing over time, and whether the trend in either direction is statistically significant.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
mk_wildfire <- Kendall::MannKendall(wildfire_ts)
summary(mk_wildfire)

#The Mann-Kendall test produces a p-value of approximately 0.0025 (<0.05) shows that it is statistically significant and there is a monotonic trend in the occurrences/area of wildfire burns over the years in California. The obtained positive tau value implies that the trend is increasing.

mk_prescribed <- Kendall::MannKendall(prescribed_fire_ts)
summary(mk_prescribed)

mk_wildland <- Kendall::MannKendall(wildland_fire_ts)
summary(mk_wildland)

#The p-value of prescribed and wildland use fires are 0.6374 and 0.5296 respectively. Hence, we can see again that these fires are not statistically significant.

```

\newpage

# Summary and Conclusions

Q1: 

Based on the p-value, we concluded that the mean sizes of large wildfires in California did not differ between 1988 and 2000. This potentially shows that 'years' would not make a good predictor for the mean sizes of large wildfires in the state, although this could have been further investigated by comparing means between different pairs of years. 

Q2:

Based on the p-value we obtained from the result of the ANOVA test, we concluded that there is a statistically significant difference in large wildland fire sizes for the states with the most large wildland fire occurrences. When running the linear model, we found that fire sizes in California and Nevada were statistically different, but Idaho was statistically the same to both Nevada and California. This was confirmed in our Tukey test (Figure 8). It would be interesting to see why Nevada and California did have different fire sizes and if it could be due to different land connectivity, land types, or management practices. These questions allow for further research and analysis. 

Q3: 

The Mann-Kendall test and time series was performed on the three categories of fires in California from 1984-2007. The Mann-Kendall test produces a p-value of approximately 0.0025 (<0.05), showing that it is statistically significant and there is a monotonic trend in the occurrences/area of wildfire burns over the years in California. The obtained positive tau value implies that the trend is increasing. The time series analysis also indicated that there is an increasing trend in the large wildfires over the years, shown in Figure 9. 

However, due to inconsistency and large gaps in the datasets of the prescribed and wildland use fires, time series analysis could not be performed on them. To further double-check this, the Mann-Kendall test was performed again, and the p-values of prescribed and wildland use fires were 0.93 and 0.68, respectively. Hence, we accept the null hypothesis and can see again that these fires are not statistically significant.


Since we know that climate change will continue to impact wildland fires globally, our analysis has allowed us to see what further questions need to be asked to get more information. Our results showed us that expanding our research to more states, years, and increased wildland fire data would allow us to see more trends and differences. With additional questions, researchers can pave the ways to further studies that can be done or gathered to understand this increase in wildfire severity and occurrences.


\newpage

# References
<add references here if relevant, otherwise delete this section> 
