---
title: "TimeSeriesAnalysis_NRS"
author: "Nanditha Ram Satagopan"
date: "`r Sys.Date()`"
output: pdf_document
---

**Time Series Analysis of Wildfires in California from 1984-2007**
Research Question:
Is there a trend in large fire occurrences in California from 1984 to 2007? What is the trend in wildfire, prescribed fires and wildland fire use over the same years? 

```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}

#Uploading necessary packages
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)

mytheme <- theme_bw(base_size = 10) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top", 
        axis.text.x=element_text(angle = 60, hjust = 1)) 
theme_set(mytheme)
```

## Wrangling datasets

```{r California Fires, echo=FALSE, warning=FALSE, message=FALSE}

#Reading in the raw dataset
Fire_data <- read.csv("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Data/Processed/US_FireData_Processed.csv",
                           stringsAsFactors = TRUE)
summary(Fire_data)

#Wrangling dataset for California fires only
all_cal_fires <- Fire_data %>%
  filter(STATE=="CA")
summary(all_cal_fires)

cal_fires <- all_cal_fires %>%
  filter(FIRE_TYPE=="Wildfire" | FIRE_TYPE=="Prescribed Fire" | FIRE_TYPE=="Wildland Fire Use")
summary(cal_fires)

#Categorizing fire datasets

wildfires_ca <- cal_fires %>%
  filter(FIRE_TYPE=="Wildfire")

prescribed_fires_ca <- cal_fires %>%
  filter(FIRE_TYPE=="Prescribed Fire")

wildland_fires_ca <- cal_fires %>%
filter(FIRE_TYPE=="Wildland Fire Use")

#Obtaining mean burn areas for each fire category

wildfires_mean <- 
  wildfires_ca %>%
  group_by(YEAR) %>%
  summarize(Mean_fires = mean(ACRES)) # %>%
  #mutate(Date = my(paste0(STARTMONTH, "-",YEAR))) 

prescribed_fires_mean <- 
  prescribed_fires_ca %>%
  group_by(YEAR, STARTMONTH) %>%
  summarize(Mean_fires = mean(ACRES))%>%
  mutate(Date = my(paste0(STARTMONTH, "-",YEAR))) 

wildland_fires_mean <- 
  wildland_fires_ca %>%
  group_by(YEAR, STARTMONTH) %>%
  summarize(Mean_fires = mean(ACRES))%>%
  mutate(Date = my(paste0(STARTMONTH, "-",YEAR))) 

```

#Exploratory Analysis

```{r exploratory analysis of TSA, echo=FALSE, warning=FALSE, message=FALSE}

#Plotting Wildfire Fires 1984-2007

wildfire_plot <- 
  ggplot(wildfires_mean, aes(x = Date, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Wildfire Area",
       y="Burnt Area in Acres",
       x="Years") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildfire_plot)

ggsave("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Output/wildfire_plot.jpg", wildfire_plot, height = 7, width = 8, units = "in", dpi = 300)

#Plotting Prescribed Fires 1984-2007

prescribed_fire_plot <- 
  ggplot(prescribed_fires_mean, aes(x = Date, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Prescribed Fire Area",
       y="Burnt Area in Acres",
       x="Years") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(prescribed_fire_plot)

ggsave("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Output/prescribed_fire_plot.jpg", prescribed_fire_plot, height = 7, width = 8, units = "in", dpi = 300)

#Plotting Wildland Fires 1984-2007

wildland_fire_plot <-
  ggplot(wildland_fires_mean, aes(x = Date, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Burnt Wildland Use Fire Area",
       y="Burnt Area in Acres",
       x="Years")+
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildland_fire_plot)

ggsave("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Output/wildland_fire_plot.jpg", wildland_fire_plot, height = 7, width = 8, units = "in", dpi = 300)

```

## Time Series Analysis

The time series analysis of the occurences of large wildfires helps us infer if they is any seasonality and/or trend in the variables over time (here, from years 1984-2007).

```{r time series, echo=FALSE, warning=FALSE, message=FALSE}

#Wildfire Time Series
wildfire_ts <- ts(wildfires_mean$Mean_fires, start = c(1984), frequency = 1)
plot(wildfire_ts)
wildfire_decomp <- stl(wildfire_ts,s.window = "periodic")
plot(wildfire_decomp)

#Prescribed Fire Time Series
prescribed_fire_ts <- ts(prescribed_fires_mean$Mean_fires, start = c(1984,1), frequency = 12)
plot(prescribed_fire_ts)
prescribed_decomp <- stl(prescribed_fire_ts,s.window = "periodic")
plot(prescribed_decomp)

#Wildland Fire Time Series
wildland_fire_ts <- ts(wildland_fires_mean$Mean_fires, start = c(1984,1), frequency = 12)
plot(wildland_fire_ts)
wildland_decomp <- stl(wildland_fire_ts,s.window = "periodic")
plot(wildland_decomp)

##The Prescribed and Wildland Fire time series are not processable since they are not periodic, or have less than two periods. Hence, we can conclude that there is no trend or pattern in these fires and hence are not significant.
```

**Mann-Kendall Test**

Running the Mann-Kendall Test to assess whether a set of data values is increasing over time or decreasing over time, and whether the trend in either direction is statistically significant.

```{r Mann-Kendall test, echo=FALSE, warning=FALSE, message=FALSE}


mk_wildfire <- Kendall::MannKendall(wildfire_ts)
summary(mk_wildfire)

#The Mann-Kendall test produces a p-value of approximately 0.0025 (<0.05) shows that it is statistically significant and there is an increasing trend in the occurrences/area of wildfire burns over the years in California.

mk_prescribed <- Kendall::MannKendall(prescribed_fire_ts)
summary(mk_prescribed)

mk_wildland <- Kendall::MannKendall(wildland_fire_ts)
summary(mk_wildland)

#The p-value of prescribed and wildland use fires are 0.93 and 0.68 respectively. Hence, we can see again that these fires are not stastically significant.
```
