---
title: "TimeSeriesAnalysis_NRS"
author: "Nanditha Ram Satagopan"
date: "`r Sys.Date()`"
output: pdf_document
---

**Time Series Analysis of Wildfires in California from 1984-2007**
Research Question:
Is there a trend in large fire occurrences in California from 1984 to 2007? What is the trend in wildfire, prescribed fires and wildland fire use over the same years? 

```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}

#Uploading necessary packages
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)

mytheme <- theme_bw(base_size = 10) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top", 
        axis.text.x=element_text(angle = 60, hjust = 1)) 
theme_set(mytheme)
```

## Wrangling datasets

```{r California Fires, echo=FALSE, warning=FALSE, message=FALSE}

#Reading in the raw dataset
Fire_data <- read.csv("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Data/Processed/US_FireData_Processed.csv",
                           stringsAsFactors = TRUE)
summary(Fire_data)

#Wrangling dataset for California fires only
all_cal_fires <- Fire_data %>%
  filter(STATE=="CA")
summary(all_cal_fires)

cal_fires <- all_cal_fires %>%
  filter(FIRE_TYPE=="Wildfire" | FIRE_TYPE=="Prescribed Fire" | FIRE_TYPE=="Wildland Fire Use")
summary(cal_fires)

#Categorizing fire datasets

wildfires_ca <- cal_fires %>%
  filter(FIRE_TYPE=="Wildfire")

prescribed_fires_ca <- cal_fires %>%
  filter(FIRE_TYPE=="Prescribed Fire")

wildland_fires_ca <- cal_fires %>%
filter(FIRE_TYPE=="Wildland Fire Use")

#Obtaining mean burn areas for each fire category

wildfires_mean <- 
  wildfires_ca %>%
  group_by(YEAR) %>%
  summarize(Mean_fires = mean(ACRES))

prescribed_fires_mean <- 
  prescribed_fires_ca %>%
  group_by(YEAR) %>%
  summarize(Mean_fires = mean(ACRES)) 

wildland_fires_mean <- 
  wildland_fires_ca %>%
  group_by(YEAR) %>%
  summarize(Mean_fires = mean(ACRES))

```

#Exploratory Analysis

```{r exploratory analysis of TSA, echo=FALSE, warning=FALSE, message=FALSE}

#Plotting Wildfire Fires 1984-2007

wildfire_plot <- 
  ggplot(wildfires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Wildfire Area",
       y="Burnt Area in Acres",
       x="Years") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildfire_plot)

ggsave("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Output/Fig5_Nanditha.jpg", wildfire_plot, height = 7, width = 8, units = "in", dpi = 300)

#Plotting Prescribed Fires 1984-2007

prescribed_fire_plot <- 
  ggplot(prescribed_fires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Prescribed Fire Area",
       y="Burnt Area in Acres",
       x="Years") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(prescribed_fire_plot)

ggsave("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Output/Fig6_Nanditha.jpg", prescribed_fire_plot, height = 7, width = 8, units = "in", dpi = 300)

#Plotting Wildland Fires 1984-2007

wildland_fire_plot <-
  ggplot(wildland_fires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Burnt Wildland Use Fire Area",
       y="Burnt Area in Acres",
       x="Years")+
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildland_fire_plot)

ggsave("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Output/Fig7_Nanditha.jpg", wildland_fire_plot, height = 7, width = 8, units = "in", dpi = 300)

##We reject the the Prescribed and Wildland Fire data sets. Their time series cannot be plotted since there are huge gaps and great inconsistency in data and thus a significant trend or pattern cannot be deduced. 

```

## Time Series Analysis

The time series analysis of the occurences of large wildfires helps us infer if they is any seasonality and/or trend in the variables over time (here, from years 1984-2007).

```{r time series, echo=FALSE, warning=FALSE, message=FALSE}

#Wildfire Time Series
wildfire_ts <- ts(wildfires_mean$Mean_fires, start = c(1984), frequency = 1)
plot(wildfire_ts)

wildfire_TS_plot <- 
  ggplot(wildfires_mean, aes(x = YEAR, y = Mean_fires)) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Wildfire Time Series",
       y="Burnt Area in Acres",
       x="Years 1984-2007") +
  theme(plot.title = element_text(size = 20))+
  theme(plot.title = element_text(hjust = 0.5))

print(wildfire_TS_plot)

ggsave("C:/Users/Admin/Desktop/Courses/Sem 1/Environmental Data Analytics/EDAProject_Go_Merritt_Satagopan/Output/Fig9_Nanditha.jpg", wildfire_TS_plot, height = 7, width = 8, units = "in", dpi = 300)
```

**Mann-Kendall Test- Trend Analysis**

Running the Mann-Kendall Test to assess whether a set of data values is increasing over time or decreasing over time, and whether the trend in either direction is statistically significant.

```{r Mann-Kendall test, echo=FALSE, warning=FALSE, message=FALSE}


mk_wildfire <- Kendall::MannKendall(wildfire_ts)
summary(mk_wildfire)

#The Mann-Kendall test produces a p-value of approximately 0.0025 (<0.05) shows that it is statistically significant and there is a monotonic trend in the occurrences/area of wildfire burns over the years in California. The obtained positive tau value implies that the trend is increasing.

mk_prescribed <- Kendall::MannKendall(prescribed_fire_ts)
summary(mk_prescribed)

mk_wildland <- Kendall::MannKendall(wildland_fire_ts)
summary(mk_wildland)

#The p-value of prescribed and wildland use fires are 0.93 and 0.68 respectively. Hence, we can see again that these fires are not stastically significant.
```
