---
title: "Two-Sample T Test"
author: "Li Jia Go"
date: "2022-11-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## Research Question

Do the mean sizes of large wildfires in Colorado differ between 2000 and 2018?

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)

USFireData <- read.csv("./Data/Processed/US_FireData_Processed.csv", 
                stringsAsFactors = TRUE) 

#filtering dataset for wildfires in California in years 1988 and 2000
California.Wildfires <- USFireData %>%
  filter(STATE %in% "CA") %>%
  filter(FIRE_TYPE %in% "Wildfire") %>%
  filter(YEAR %in% c("1988", "2000"))

#data exploration 
ggplot(California.Wildfires, aes(x = ACRES, color =as.factor(YEAR))) +
  geom_freqpoly() +
  labs(x='Acreage of fires', y='Count', color='Year', 
       title="Histogram of wildfires in 1988 & 2000, California")  
  

```
### Two-sample t-test

The two-sample *t* test is used to test the hypothesis that the mean of two samples is equivalent. Unlike the one-sample tests, a two-sample test requires a second assumption that the variance of the two groups is equivalent. 
H0: means are equal
h1: means are not equal. 

For example, we might ask ourselves (from the EPA air quality processed dataset): 
Are Ozone levels different between 2018 and 2019?

```{r}
#First let's look at the data
ggplot(EPAair, aes(x = Ozone, color = as.factor(Year))) +
  geom_freqpoly()
#hard to tell if mean is the same for both plots


#Format as a t-test
#EPAair$Ozone will be our continuous dependent variable
#EPAair$Year will be our categorical variable with two levels (2018 and 2019)
O3.twosample <- t.test(EPAair$Ozone ~ EPAair$Year)
O3.twosample
#p <0.05, reject null, accept alternative that 2 means are not equal. alternative hypothesis: true difference in means between group 2018 and group 2019 is not equal to 0


RawUSFireData <- read.csv("./Data/Raw/US_FireData_Raw.csv", 
                stringsAsFactors = TRUE) 
# Format as a GLM
O3.twosample2 <- lm(EPAair$Ozone ~ EPAair$Year) #ask if year is predictor of ozone level ? 
summary(O3.twosample2)
#for EPAair$year: p value <0.05, year is statistically different from 0 and is a predictor of ozone level

