---
title: "Two-Sample T Test"
author: "Li Jia Go"
date: "2022-11-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## Research Question

Do the mean sizes of large wildfires in California differ between 1988 and 2000?

```{r, data exploration}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(sf)
library(mapview)


Fire_data <- read.csv("./Data/Processed/US_FireData_Processed.csv") 

#filtering dataset for wildfires in California in years 1988 and 2000
California.Wildfires <- Fire_data %>%
  filter(STATE %in% "CA") %>%
  filter(FIRE_TYPE %in% "Wildfire") %>%
  filter(YEAR %in% c("1988", "2000"))
```

## Exploratory Analysis 
```{r, exploratory analysis}
#data exploration

California_hist_1988_2000 <- ggplot(California.Wildfires, aes(x = ACRES, color =as.factor(YEAR))) +
  geom_freqpoly() +
  labs(x='Acreage of fires', y='Count', color='Year', 
       title="Histogram of wildfires in 1988 & 2000, California")
print(California_hist_1988_2000)

ggsave("./Output/Fig1_Li.jpg",California_hist_1988_2000 , height = 7, width = 8, units = "in", dpi = 300)

#exploratory spatial plots 
California.Wildfires.sf <- California.Wildfires %>% 
  st_as_sf(coords=c('LONGITUDE', 'LATITUDE'), crs=4269)

Cali.Wildfire.Plot = mapview(California.Wildfires.sf, zcol="YEAR")
Cali.Wildfire.Plot




```
### Two-sample t-test

The two-sample *t* test is used to test the hypothesis that the mean of two samples is equivalent. Unlike the one-sample tests, a two-sample test requires a second assumption that the variance of the two groups is equivalent. 
H0: means are equal
h1: means are not equal. 


```{r, two-sample t test}

#running two-sample t test
twosamp.ttest <- t.test(California.Wildfires$ACRES ~ California.Wildfires$YEAR)
twosamp.ttest
#p= 0.3754>0.05, we accept the null hypothesis that means are statistically equal for both years. 


```